---
title: "Electricity generation by countries"
author:
  - name: "Eric Bidong Wakunga"
    orcid: "0009-0001-3770-5405"
    email: "eric.bidongsr@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: html
execute: 
  warning: false
editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: apa-5th-edition.csl
link-citations: true
---

# Introduction

This dataset [@ourworldindata2025electricity] provides historical electricity generation data TWh for countries and regions around the world, covering the year 1985 through 2024. This data was compiled to track energy production trends over time and support energy planning, policy evaluation, and investment analysis. By analyzing this dataset, we can identify which countries have scaled their electricity systems significantly and which ones have stagnated. This analysis is particularly relevant for global efforts toward energy access, sustainability, and emissions reduction.

# Methods

The electricity generation dataset was compiled by *Our World in Data* based on original sources from Ember and the Energy Institute's *Statistical Review of World Energy (2025)*. It includes major post-processing by Our World in Data to standardize and harmonize national statistics. The dataset covers total electricity generated per country annually, measured in terawatt-hours, over the period 1985 to 2024. It was last updated in June 2025 and is intended to support global energy analysis and policy research.

## **Loading libraries**

```{r}
#import libraries

library(tidyverse)
library(here)
library(dplyr)
library(RColorBrewer)
library(ggplot2)
library(gt)
library(ggrepel)
```

## Reading data

```{r}
energy_generation_by_countries <- read_csv(here("data/raw/capston_dataset.csv"))
```

## Data Exploratory Approach

### Data summary

```{r}
dim(energy_generation_by_countries)
```

This dataset has 7653 rows and 4 columns.

Let's visualize the first five rows to inspect the data.

```{r}

head(energy_generation_by_countries)
```

The data looks good. There are some countries that are missing Code, but this is not important since we have the names of these countries.

The data types are also correct.

Let's group data by country and summarize the statistics for exploration.

```{r}

energy_summary <- energy_generation_by_countries |>
  group_by(Entity) |>
  summarise(
    total_generation_TWh = sum(`Electricity generation - TWh`, na.rm = TRUE),
    n_years = n()
  ) |>
  arrange(desc(total_generation_TWh))

energy_summary
```

As you can see, there are aggregated data by region (OECD, High-income countries, etc) that are included in this dataset. In the next section, we will remove those in order to stick with only countries.

Let's now see the electricity generation growth rate by countries

```{r}
growth_rates <- energy_generation_by_countries |>
  arrange(Entity, Year) |>
  group_by(Entity) |>
  mutate(
    prev_year_generation = lag(`Electricity generation - TWh`),
    annual_growth_rate = (`Electricity generation - TWh` - prev_year_generation) / prev_year_generation
  ) |>
  summarise(
    total_generation_TWh = sum(`Electricity generation - TWh`, na.rm = TRUE),
    avg_annual_growth = mean(annual_growth_rate, na.rm = TRUE),
    min_year = min(Year),
    max_year = max(Year),
    n_years = n() - 1
  ) |>
  arrange(desc(avg_annual_growth))

growth_rates
```

Energy generation growth rate is high in developing countries. This shows that much effort is put recently to increase the generation capacity. We could interestingly compare such data with GDP data or any other wealth metric to understand better what is going on in each country.

## **Initial data Tidying**

### Removing aggregated data by region

As shown previously, our dataset includes aggregated data which can bias our analysis. So we have to remove them.

```{r}
entity_to_remove <- c(
  "World", "G20 (Ember)", "High-income countries", "OECD (EI)", "Non-OECD (EI)", "Asia", "OECD (Ember)", "Asia Pacific (EI)", "Upper-middle-income countries", "Asia (Ember)", 
"G7 (Ember)", "North America (EI)", "North America", "Europe", "Europe (EI)", "Europe (Ember)", "North America (Ember)", "European Union (27)","EU (Ember)","Lower-middle-income countries","CIS (EI)", "South and Central America (EI)", "Latin America and Caribbean (Ember)","South America", "Middle East (EI)", "Middle East (Ember)", "Africa", "Africa (EI)", "ASEAN (Ember)", "Africa (Ember)", "Oceania", "Oceania (Ember)", "Eastern Africa (EI)","Western Africa (EI)", "Central America (EI)", "Middle Africa (EI)", 'Western Sahara'
)

energy_generation_by_countries <- energy_generation_by_countries |>
  filter(!Entity %in% entity_to_remove)
```

Now the aggregates have been remove, let's recompile the energy_summary by countries and update the dataframe.

```{r}
energy_summary <-energy_generation_by_countries |>
  arrange(Entity, Year) |>
  group_by(Entity) |>
  mutate(
    prev_year_generation = lag(`Electricity generation - TWh`),
    annual_growth_rate = (`Electricity generation - TWh` - prev_year_generation) / prev_year_generation
  ) |>
  summarise(
    total_generation_TWh = sum(`Electricity generation - TWh`, na.rm = TRUE),
    avg_annual_growth = mean(annual_growth_rate, na.rm = TRUE),
    min_year = min(Year),
    max_year = max(Year),
    n_years = n() - 1
  ) |>
  arrange(desc(total_generation_TWh))

energy_summary
```

### Renaming columns

Let lower the names of the columns and remove all the spaces

```{r}
energy_generation_by_countries <- energy_generation_by_countries |>
  rename_with(~ str_to_lower(.)) |>
  rename_with(~ str_replace_all(., "[\\s\\-]+", "_")) |>
  rename_with(~ str_replace_all(., "[^a-z0-9_]", ""))

energy_generation_by_countries
```

### Saving data

```{r}
write_csv(energy_generation_by_countries, here("data/processed/energy_production.csv"))
```

# Results

## Time coverage in data by each country

We want to check for what period of time there is electrical generation data for each country before performing other EDA.

```{r}
# Create a new label column showing "start–end" range
energy_summary_labeled <- energy_summary %>%
  mutate(range_label = paste(min_year, max_year, sep = "–"))
```

```{r}
#| label: fig-electricity-eneration-data-time-span
#| fig-cap: Electricity Generation Data Time Span.
#| alt: Bar plot with year spans representing the availability of data.

ggplot(energy_summary_labeled, aes(y = reorder(range_label, min_year),  
                                   x = min_year,
                                   xend = max_year)) +
  geom_segment(aes(xend = max_year, yend = range_label),
               size = 2, color = "darkgreen") +
  labs(
    title = "Electricity Generation Data Time Span",
    x = "Year",
    y = "Time Range (Start–End Year)"
  ) +
  scale_x_continuous(breaks = seq(1985,2024, by=2), limits = c(1985,2024))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Not all the countries have data for the 39-year period. the @fig-electricity-eneration-data-time-span graph provides the span for each category. This will help to better understand trend later in our analysis.

## Average annual growth vs total generation

Let's look at the relationship between electricity generation growth vs total generation.

```{r}
#| label: fig-growth-vs-total-annotated
#| fig-cap: Electricity generation vs growth rate.
#| alt: Scatter plot with labeled bubbles showing countries’ total electricity generation and growth rate. Labels avoid overlap using leader lines.

#Create a variable called growth_percent that converts annual growt in % format
energy_annotated <- energy_summary %>%
  mutate(growth_percent = avg_annual_growth * 100)  # Convert to percent

ggplot(energy_annotated, aes(x = total_generation_TWh, y = growth_percent)) +
  geom_point(color = "darkorange", size = 3, alpha = 0.7) +                      # Bubbles
  geom_smooth(method = "lm", se = FALSE, color = "gray40", linetype = "dashed") +  # Trendline
  geom_text_repel(aes(label = Entity), size = 3, max.overlaps = 100) +          # Non-overlapping country labels
  labs(
    title = "Electricity Generation vs Growth Rate (with Country Labels)",
    x = "Total Generation (TWh)",
    y = "Average Annual Growth Rate (%)"
  ) +
  theme_minimal()
```

From this @fig-growth-vs-total-annotated, it turns out that industrialized countries such USA, China, Japon, India, South Corea have low growth rate compared to developing countries. China generation capacity is growing faster that USA, yet they generation capacity is close to each other. Palestine has the biggest growth rate, though it generation still is low.

Let's group countries by low midium, high and very high growth rate in the @tbl-growth-class-list

```{r}
#| label: tbl-growth-class-list
#| tbl-cap: Grouping of countries by average annual electricity generation growth rate.


# Step 1: Classify growth and assign sorting order
classified_growth <- energy_summary %>%
  mutate(
    growth_class = case_when(
      avg_annual_growth <= 0.02 ~ "Low (≤ 2%)",
      avg_annual_growth <= 0.10 ~ "Medium (>2% – ≤10%)",
      avg_annual_growth <= 0.15 ~ "High (>10% – ≤15%)",
      avg_annual_growth >  0.15 ~ "Very High (> 15%)"
    ),
    class_order = case_when(
      growth_class == "Low (≤ 2%)" ~ 4,
      growth_class == "Medium (>2% – ≤10%)" ~ 3,
      growth_class == "High (>10% – ≤15%)" ~ 2,
      growth_class == "Very High (> 15%)" ~ 1
    )
  )

# Step 2: Group and list countries per class
growth_summary <- classified_growth %>%
  group_by(growth_class, class_order) %>%
  summarise(`List of Countries` = paste(sort(Entity), collapse = ", "), .groups = "drop") %>%
  arrange(class_order) %>%
  select(`Growth Category` = growth_class, `List of Countries`)

# Step 3: Render the table
gt(growth_summary) %>%
  tab_header(title = "Countries by Growth Rate Category") %>%
  cols_width(`List of Countries` ~ px(600))
```

## Generation trend

Let's look at the historical trend and see who which countries are showing positive pace.

```{r}
#| label: fig-electricity-eneration-trend
#| fig-cap: Electricity Generation trend for each countries
#| alt: line plot representing trend in the electricity generation.

# Step 1: Identify top 15 countries by total generation
top_15 <- energy_summary |>
  arrange(desc(total_generation_TWh)) |>
  slice(1:15) |>
  pull(Entity)

# Step 2: Add color flag to the full dataset
energy_trend_labeled <- energy_generation_by_countries |>
  rename(generation_TWh = electricity_generation_twh) |>
  mutate(highlight = ifelse(entity %in% top_15, entity, "Other"))

# Step 3: Define distinct color palette for top 15 + gray for others
top_colors <- setNames(c(brewer.pal(12, "Set3"), brewer.pal(3, "Dark2")), top_15)
all_colors <- c("Other" = "gray80", top_colors)

# Step 4: Plot
ggplot(energy_trend_labeled, aes(x = year, y = generation_TWh, group = entity, color = highlight)) +
  geom_line(alpha = 0.8, linewidth = 0.7) +
  scale_color_manual(
    values = all_colors,
    breaks = top_15,                     # legend will only show the top 15
    guide = guide_legend(override.aes = list(linewidth = 1.5))
  ) +
  labs(
    title = "Electricity Generation Trend by Country",
    x = "Year",
    y = "Electricity Generation (TWh)",
    color = "Top 15 Countries"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```

This @fig-electricity-eneration-trend graph show how China growth is significant, it is far way higher than USA that he crossed in around 2010. He is followed by USA which the trend seem flat and India which has taken off. Other developed countried like Russia, France, Japan, etc. also have a flat trend, no visible growth.

## Summary statistics of China

As the chart above has depicted Chine and USA, let see what their summary statistics are in the table @tbl-china-usa-summary below.

```{r}
#| label: tbl-china-usa-summary
#| tbl-cap: Summary statistics of electricity generation for China and the USA


# Filter for China and USA
china_usa_summary <- energy_generation_by_countries %>%
  filter(entity %in% c("China", "United States")) %>%
  group_by(entity) %>%
  summarise(
    Start = min(year, na.rm = TRUE),
    End = max(year, na.rm = TRUE),
    `Nb Year` = n(),
    `Total generation [TWh]` = sum(`electricity_generation_twh`, na.rm = TRUE),
    `Avg generation [TWh]` = mean(`electricity_generation_twh`, na.rm = TRUE),
    `Min generation [TWh]` = min(`electricity_generation_twh`, na.rm = TRUE),
    `Max generation [TWh]` = max(`electricity_generation_twh`, na.rm = TRUE),
    `Avg annual growth` = (last(`electricity_generation_twh`) / first(`electricity_generation_twh`))^(1/(`Nb Year` - 1)) - 1)

gt(china_usa_summary) %>%
  fmt_percent(columns = `Avg annual growth`, decimals = 2) %>%
  fmt_number(columns = matches("Generation"), decimals = 1)
```

# Conclusions

The analysis of historical electricity generation data from 1985 to 2024 reveals significant differences in both the scale and growth of energy production across the 215 countries. High-output nations such as China and the United States dominate in total generation, while several smaller economies demonstrate rapid growth path. By classifying countries into growth categories, we can distinguish between mature, expanding, and lagging electricity systems which offers insights for policymakers, investors, and energy planners. The dataset also highlights the uneven availability of historical data, which can affect some types of comparisons. Overall, this dataset provides a valuable foundation for understanding global electricity trends and important strategic decisions toward energy transition and access.

## Summary of Findings

This project provide a glimpse to undertand the energy generation trend around the globe. The analysis is conducted on 215 countries and yield the following result:

1.  China is leading the energy generation race globally, along with many of the developping countries such as Benin and Palestine.

2.  India also showing a wake up trend.

3.  With the current trend, China generation capacity will surpass the USA, as it has done for Japan, Russia, and other developed countries.

### Questions and Next Steps

1.  How does electricity generation correlate with development indicators
2.  Are countries with low or negative growth facing structural challenges?
3.  What is the role of renewable energy in total generation and growth patterns
